---
title: "UMBS LTREB Saplings Table Cleaning"
author: "Alexandria Pawlik"
output: html_notebook
---

This is an R notebook detailing the cleaning process for LTREB saplings 2014 data

```{r}
library(tidyverse)
library(dataMaid)
library(lubridate)
```

Read in csv file
```{r}
saplings <- read_csv("../raw/corrected sapling tally and biomass for Alina.csv")
saplings <- saplings %>% select(1:16,18:25,27)
head(saplings)
```


Clean up column headers
```{r}
saplings <- saplings %>% rename("DBH_0_to_2" = `0-2`, "DBH_0_to_2_dead" = `0-2, Dead`, "DBH_2_to_4" = `2-4`, "DBH_2_to_4_dead" = `2-4, dead`, "DBH_4_to_6" = `4-6`, "DBH_4_to_6_dead" = `4-6, dead`, "DBH_6_to_8" = `6-8`, "DBH_6_to_8_dead" = `6-8, dead`, "Biomass_0_to_2" = `Biomass 0-2`, "Biomass_0_to_2_dead" = `Biomass 0-2, dead`, "Biomass_2_to_4" = `Biomass 2-4`, "Biomass_2_to_4_dead" = `Biomass 2-4, dead`, "Biomass_4_to_6" = `Biomass 4-6`, "Biomass_4_to_6_dead" = `Biomass 4-6, dead`, "Biomass_6_to_8" = `Biomass 6-8`, "Biomass_6_to_8_dead" = `Biomass 6-8, dead`, "Mean_height_0_to_2" =`0-2 Mean Height`, "Mean_height_2_to_4" =`2-4 Mean Height`, "Mean_height_4_to_6" =`4-6 Mean Height`, "Mean_height_6_to_8" =`6-8 Mean Height`)

saplings$"Mean_height_0_to_2_dead" = saplings$Mean_height_0_to_2
saplings$"Mean_height_2_to_4_dead" = saplings$Mean_height_2_to_4
saplings$"Mean_height_4_to_6_dead" = saplings$Mean_height_4_to_6
saplings$"Mean_height_6_to_8_dead" = saplings$Mean_height_6_to_8

head(saplings)
```

Fix up tally date column
```{r}
saplings$`Tally Date` <- mdy(saplings$`Tally Date`)
head(saplings)
#glimpse(saplings)
```



Merge/gather columns to have "size class" "tally" "biomass" columns, ie wide to long
Done by splitting into two data frames (tally and biomass), gather and reorder separately, and then bind necessary columns

```{r}
saplings_dbh <- saplings %>% select(1:5,7,8,10,11,13,14,15,25)
saplings_bio <- saplings %>% select(1,2,3,15, 17:25)
saplings_mean <- saplings %>% select(1:3, 6, 26, 9, 27, 12, 28, 16, 29, 15, 25)

saplings_dbh <- saplings_dbh %>% gather(DBH_size_class, Tally, -Plot, -Species, -`Tally Date`, -`Total Biomass (kg)`, -Index) %>% arrange(Index)

saplings_bio <- saplings_bio %>% gather(DBH_size_class, Biomass, -Plot, -Species, -`Tally Date`, -`Total Biomass (kg)`, -Index) %>% arrange(Index)

saplings_mean <- saplings_mean %>% gather(DBH_size_class, Mean_Height, -Plot, -Species, -`Tally Date`, -`Total Biomass (kg)`, -Index) %>% arrange(Index)

head(saplings_mean)
```

Edit out "DBH" and "Biomass" from size class column
```{r}
saplings_dbh <- saplings_dbh %>% mutate(DBH_size_class = str_sub(DBH_size_class, 5, -1))
saplings_bio <- saplings_bio %>% mutate(DBH_size_class = str_sub(DBH_size_class, 9, -1))
saplings_mean <- saplings_mean %>% mutate(DBH_size_class = str_sub(DBH_size_class, 13, -1))


glimpse(saplings_dbh)
glimpse(saplings_bio)
glimpse(saplings_mean)
```

Merge and reorder
```{r}
saplings_tidy <- cbind(saplings_dbh, Biomass = saplings_bio[[7]], Mean_height = saplings_mean[[7]])
head(saplings_tidy)
saplings_tidy <- saplings_tidy %>%  select(2:4, 6:9, 5)
glimpse(saplings_tidy)
```

Mutate size class column into size class and status
```{r}
saplings_tidy <- saplings_tidy %>% mutate(Status = ifelse(str_sub(DBH_size_class, -1, -1) =="d", "Dead", "Alive"))
head(saplings_tidy)
saplings_tidy <- saplings_tidy %>% mutate(DBH_size_class = str_sub(DBH_size_class, 1, 6)) %>% select(1:4, 9, 5:7)

```

fix tally, change NA to 0
```{r}
saplings_tidy$Tally[is.na(saplings_tidy$Tally)] <- 0
saplings_tidy$Biomass[is.na(saplings_tidy$Biomass)] <- 0
```

rename columns
```{r}
saplings_tidy <- saplings_tidy %>% rename("Tally_date" = `Tally Date`) %>% select(1:7)
```

fix plot names to match AGB
```{r}
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#1"] <- "DNR1"
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#2"] <- "DNR2"
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#3"] <- "DNR3"
plot_names <- unique(saplings_tidy$Plot)
plot_names
```



Write output file
```{r}
write_csv(saplings_tidy, path = "../output/UMBS_Saplings_Inventory_2014_tidy.csv")
```




