---
title: "UMBS LTREB Above Ground Biomass Table Cleaning"
author: "Alexandria Pawlik"
output: html_notebook
---

Required packages and paths
```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(xlsx)
library(tibble)
library(dataMaid)
options(stringsAsFactors = FALSE)

# set these paths for your machine's specifics
master_path_2014 <- "../raw/2014 exported/Master AGB data for UMBS IMS merged master tree list.csv"
live_path_2014 <- "../raw/2014 exported/live inventory1.csv"
dead_path_2014 <- "../raw/2014 exported/dead inventory.csv"
clean_2014_path <- "../edi.243.2/UMBS_LTREB_AboveGroundBiomass.csv"
raw_2019_path <- "../raw/AGB_2019_raw.csv"
output_path <- "../output/agb.csv"
```

*Note: This notebook describes the merging process between the Master AGB data and the 2014 tree inventory. Currently only includes Live and Dead and excludes Sapling*



Read in merged master data (pulled from "Master AGB data for UMBS IMS.xlsx" tab "merged master tree list" as a csv file)
```{r}
master_data <- read_csv(master_path_2014)
glimpse(master_data)
```

Read in Live OS data (pulled from "Tree Inventory - Master.xlsx" tab "Live OS Inventory" as a csv file)
```{r}
live_data <- read_csv(live_path_2014)
glimpse(live_data)
```

Read in Dead OS data (pulled from "Tree Inventory - Master.xlsx" tab "Dead OS Inventory" as a csv file)
```{r}
dead_data <- read_csv(dead_path_2014)
#glimpse(dead_data)
```

Mapping between species codes and common names, because there are no NA values for species codes but inconsistent entry for common names, allows for common names to be uniform

**Naming anomalies based on Permanent Plots data  
SASP is technically SALI, for Salix spp.  
VISP is technically VIBU for Viburnum spp.  
PYSP is technically PY for Pyrus  
OTHER implies a species not in the list below was measured  

**How do we want to deal with Unknowns? Is "OTHER" considered unknown?   
```{r}
index <- c("POGR", "PIRE", "PIST", "ACRU", "QURU", "TSCA", "BEPA", "FAGR", "AMSP", "ACSA", "ACPE", "FRAM", "OSVI", "TIAM", "ULAM", "UNKNOWN", "ABBA", "POTR", "THOC", "PIGL", "PRSE", "PRPE", "COAL", "RHGL", "CRRO", "FRNI", "POBA", "ALIN", "LALA", "BEAL", "SASP", "VISP", "PYSP", "PIMA", "OTHER")
values <- c("Populus grandidentata", "Pinus resinosa", "Pinus strobus", "Acer rubrum", "Quercus rubra", "Tsuga canadensis", "Betula papyrifera", "Fagus grandifolia", "Amelanchier", "Acer saccharum", "Acer pensylvanicum", "Fraxinus americana", "Ostrya virginiana", "Tilia americana","Ulmus americana", "UNKNOWN", "Abies balsamea", "Populus tremuloides", "Thuja occidentalis", "Picea glauca", "Prunus serotina", "Prunus pensylvanica", "Cornus alternifolia", "Rhus glabra", "Crataegus roanensis", "Fraxinus nigra", "Populus balsamifera", "Alnus incana", "Larix laricina", "Betula lutea", "Salix spp.", "Viburnum spp.", "Pyrus", "Picea mariana", "OTHER")

live_data <- add_column(live_data, "Common_name" = NA, .after = 8)
live_data$'Common_name' <- values[match(live_data$Species, index)]
```

```{r}
dead_data <- add_column(dead_data, "Common_name" = NA, .after = 8)
dead_data$'Common_name' <- values[match(dead_data$Species, index)]
```

Sort live and dead data sets so that the columns match better

** Assumptions being made when modifying Live and Dead data:  
1) Tag # = LTREB tag  
2) Date should be converted to 2014 for Live and Dead (ignore month/day)  
3) DBH_inches should probably be left blank unless converted to 1in:2.54cm from DBH_cm
4) Biomass estimate (kg) = AGB_kg  
5) Height (estimate) = ht (height is an estimate of tree with broken top in Dead data?)  
6) Ignore the following columns: Outside historic plot, distance, azimuth, unit circle, unit circle radians, status, notes, plot center x, plot center y, tree x, tree y, random number, volume estimate, biomass estimate (g)

** To consider:  
- Origin_year and Stand_year are both empty, can potentially be extracted from plot or removed  
- Do we want to include status from 2014 inventory? (Dominant (D), Co-dominant (C), Overstory (O), Dead (X), etc)  
- Do we want to convert cm to inches for DBH? Keep both or just one?  
- How do we want to deal with dates? Just assign 2014 to live and dead, ignore m/d?  
- Are there any more columns that should be included/excluded?  
```{r}
live_data_sorted <- live_data 
live_data_sorted <- live_data_sorted[c(3, 9, 8, 13, 1, 11, 20, 12, 2, 4, 5, 6, 7, 10, 14, 15, 16, 17, 18, 19)]
glimpse(live_data_sorted)
```

```{r}
dead_data_sorted <- dead_data
dead_data_sorted <- dead_data_sorted[c(3, 9, 8, 13, 1, 11, 22, 12, 2, 4, 5, 6, 7, 10, 14, 15, 16, 17, 18, 19, 20, 21)]
```

Add in missing columns 

```{r}
live_data_sorted <- add_column(live_data_sorted, "Measure_year" = 2014, .after = 4)
live_data_sorted <- add_column(live_data_sorted, "Sort" = NA, .before = 1) 
live_data_sorted <- add_column(live_data_sorted, "Old_tag" = NA, .before = 7)
live_data_sorted <- add_column(live_data_sorted, "Data_set" = "Live", .after = 13)
glimpse(live_data_sorted)
```

```{r}
dead_data_sorted <- add_column(dead_data_sorted, "Measure_year" = 2014, .after = 4)
dead_data_sorted <- add_column(dead_data_sorted, "Sort" = NA, .before = 1)
dead_data_sorted <- add_column(dead_data_sorted, "Old_tag" = NA, .before = 7)
dead_data_sorted <- add_column(dead_data_sorted, "Data_set" = "Dead", .after = 13)
dead_data_sorted <- add_column(dead_data_sorted, "Outside Historic Plot" = NA, .before = 7)

#glimpse(dead_data_sorted)
```


Temporary data sets that will be combined into the master set (step to take out unwanted columns, prepare Live and Dead for merging)   
** Added in temporary Data Set column to help debugging process, denotes which data set a row came from 
```{r}
live_to_combine <- live_data_sorted
head(live_to_combine)
live_to_combine <- live_to_combine[c(1:15, 18,19)]
live_to_combine <- live_to_combine %>% rename("Species_code" = Species, "LTREB_tag" = `Tag #`, "AGB_kg" = `Biomass estimate (kg)`, "Height_dead" = `Height (estimate)`, "DBH_cm" = `DBH (cm)`, "Distance" = `Distance (m)` )
glimpse(live_to_combine)
```

```{r}
dead_to_combine <- dead_data_sorted
#glimpse(dead_to_combine)
dead_to_combine <- dead_to_combine[c(1:12, 14:16, 19,20)]
#glimpse(dead_to_combine)
dead_to_combine <- dead_to_combine %>% rename("Species_code" = Species, "LTREB_tag" = `Tag #`, "AGB_kg" = `Biomass estimate (kg)`, "Height_dead" = `Height (m) estimate`, "DBH_cm" = `DBH (cm)`, "Distance" = `Distance (m)` )

#glimpse(dead_to_combine)
```

```{r}
master_to_combine <- master_data
glimpse(master_to_combine)
master_to_combine <- master_to_combine[c(1,2,4,5,6,7,8, 11,12,13)]
glimpse(master_to_combine)
master_to_combine <- add_column(master_to_combine, "Data_set" = "Master", .after = 10)
master_to_combine <- rename(master_to_combine,"Species_code" = `Species code`, "Old_tag" = `old tag`, "LTREB_tag" = `LTREB tag`, "DBH_cm" = `dbh (cm)`, "Height_stem" = `ht`, "Common_name" = `Common Name`, "Measure_year" = `Measure Year`, "AGB_kg" = `AGB (kg)`)
glimpse(master_to_combine)
```

Combine data sets
```{r}
master_to_combine <- add_column(master_to_combine, "Status" = NA, .after = 11)
master_to_combine <- add_column(master_to_combine, "Outside Historic Plot" = NA, .after = 12) 
master_to_combine <- add_column(master_to_combine, "Notes" = NA, .after = 13) 
master_to_combine <- add_column(master_to_combine, "Distance" = NA, .after = 14)
master_to_combine <- add_column(master_to_combine, "Azimuth" = NA, .after = 15)
master_to_combine <- add_column(master_to_combine, "Date" = NA, .after = 16)
master_to_combine <- add_column(master_to_combine, "Height_dead" = NA, .after = 17)
live_to_combine <- add_column(live_to_combine, "Height_stem" = NA, .after = 17)
dead_to_combine <- add_column(dead_to_combine, "Height_stem" = NA, .after = 17)
head(master_to_combine)
head(live_to_combine)
head(dead_to_combine)

setdiff(names(master_to_combine), names(dead_to_combine))

new_master <- rbind(master_to_combine, live_to_combine, dead_to_combine)
glimpse(new_master)
```


Using species code to common names mapping, fixes the majority of common name inconsistencies (any species codes not in the mapping will have NA common name). Includes lists of unique Common_name and Species code values
```{r}
new_master$"Species_code"[new_master$"Species_code" == "UNKOWN"] <- "UNKNOWN" #fix misspellings
new_master$"Species_code"[new_master$"Species_code" == "Coal"] <- "COAL"
sc <- unique(new_master["Species_code"])
sc
new_master$'Common_name' <- values[match(new_master$"Species_code", index)]
cns <- unique(new_master["Common_name"])
cns

head(new_master)
```

Remove characters from "Old_tag" column (redundant because separate columns for tags)
```{r}
bad_names <- c("Old Tag ", "Old tag ")
new_master$"Old_tag" <- ifelse(substr(new_master$"Old_tag", 1, 8)  %in% bad_names, substr(new_master$"Old_tag", 8, str_length(new_master$"Old_tag")), new_master$"Old_tag")
new_master$"Old_tag" <- ifelse(substr(new_master$"Old_tag", 1, 7)  == "Old tag", substr(new_master$"Old_tag", 7, str_length(new_master$"Old_tag")), new_master$"Old_tag")
```


Add sort values for live and dead data (continued from where master left off)
```{r}
new_master$"Sort" <- c(1:nrow(new_master))
tail(new_master)
```

Add new column "As_class" to determine if DBH measurement was written as a "size class" (ie, size 1-10 inches) or an actual measurement. "Yes" if was a size class, "NA" if not  

**According to the Permanent Plot data, only the 50 Year plots had this kind of data. Below are details for each plot  

50yr2 - 1938, 1945, 1951, 1968, 1973, 1979 all size classes  
      - 1988, 1991, 1998, 2000, 2010 all actual measurements  

50yr3 - 1938, 1945, 1951, 1968, 1973, 1979 all size classes  
      - 1988, 1991, 1998, 2000, 2010 all actual measurements  

50yr4 - 1938, 1945, 1951, 1955, 1968, 1973, 1979 all size classes  
      - 1991, 2000, all actual measurements  

50yr5 - 1938, 1945, 1951, 1955, 1973, 1979 all size classes  
      - 1991, 2001, all actual measurements  

50yr6 - 1938, 1945, 1951, 1955, 1981, 1991 all size classes  


50yr7 - 1938, 1945, 1951, 1955, 1981 all size classes  
      - 1991 all actual measurements  
      
```{r}
new_master <- add_column(new_master, "As_class" = NA, .after = 11)
similar <- c("50YR#2", "50YR#3", "50YR#4", "50YR#5")
new_master$"As_class"[new_master$"Measure_year" <= 1979 & new_master$"Plot" %in% similar] <- "Yes"
new_master$"As_class"[new_master$"Plot" == "50YR#6"] <- "Yes"
new_master$"As_class"[new_master$"Measure_year" <= 1981 & new_master$"Plot" == "50YR#7"] <- "Yes"
head(new_master)
```



```{r}
new_master <- add_column(new_master, "Origin_year" = NA, .after = 19)
new_master <- add_column(new_master, "Stand_year" = NA, .after = 19)
```


```{r}
plot_names <- unique(new_master["Plot"])
plot_names
```


Missing origin years per plots encoding:  
1890 = BS42  
-0 = BF  
matches last 2 digits = BP  
1880 = 50YR  
1839 = CP  
1987 = DNR  
1972 = PSA 

1879 = BS34
1890 = BS40
1910 = AC
1920 = Wells
 
50YR 2-5 as listed
```{r}
plot_index <- c("BS34A", "BS40E", "BS40F", "BS42A", "BS42A&BS42C", "BS42B", "BS42C", "BS34B", "AC103", "AC104", "AC107", "Wells", "BF1", "BF2", "BP54#1", "BP54#2", "BF5", "BF6", "BP36#1", "BP36#2", "50YR#2", "50YR#3", "50YR#4", "50YR#5", "BS40A", "50YR#6", "50YR#7", "BF7", "BF8", "BP80#1", "BP80#2", "BP80#3", "BP98#1", "BP98#2", "BS40X", "CP1", "CP2", "CP3", "DNR1", "DNR2", "DNR3", "PSA1", "PSA2", "PSA3")
plot_values <- c("1879", "1890", "1890", "1890", "1890", "1890", "1890", "1879", "1910", "1910", "1910", "1920", "1954", "1954", "1954", "1954", "1936", "1936", "1936", "1936", "1908", "1919", "1921", "1919", "1712", "1880", "1880", "1952", "1952", "1980", "1980", "1980", "1998", "1998", "1890", "1839", "1839", "1839", "1987", "1987", "1987", "1972", "1972", "1972")

df <- data.frame(plot_index, plot_values)
glimpse(df)


plot_index
plot_values
```

Fill in Origin year column based on above dictionary
```{r}
new_master$'Origin_year' <- plot_values[match(new_master$Plot, plot_index)]
new_master$Origin_year <- as.numeric(new_master$Origin_year)
```

Fill in Stand year column (Measure - Origin)
```{r}
new_master$Stand_year <- new_master$Measure_year - new_master$Origin_year
```

Check origin year dictionary is correct
```{r}
for (i in 1:length(plot_values)) {
  cat(plot_index[i], "<-", plot_values[i], "\n")
}
```





Migrate old tags from notes column
```{r}
u_notes <- unique(new_master$Notes)

new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 8)=="Old tag ", str_sub(Notes, 9, -1), Old_tag))
new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 7)=="Old tag", str_sub(Notes, 8, -1), TEMP))
new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 8)=="old tag ", str_sub(Notes, 9, -1), TEMP)) #note, this makes things like "old tag unreadable" into "unreadable" in Old_tag column, but maybe thats ok? easy to remove non-numerics later

new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 3)=="OT#", str_sub(Notes, 4, -1), TEMP))

new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 9)=="Old Tag #", str_sub(Notes, 10, -1), TEMP))
new_master <- mutate(new_master, TEMP = ifelse(str_sub(Notes, 1, 4)=="Tag ", str_sub(Notes, 5, -1), TEMP))

# new_master$Old_tag[new_master$Old_tag == "112 - KJN - BS42B"] <- "112" #do we want this?
# new_master$Old_tag[new_master$Old_tag == "23 - KJN - BS42B"] <- "23" #do we want this?
```


Merge over from TEMP column, get rid of OT# in Old_tag column
```{r}
new_master$Old_tag <- ifelse(is.na(new_master$Old_tag) & !is.na(new_master$TEMP), new_master$TEMP, new_master$Old_tag)


new_master <- mutate(new_master, Old_tag = ifelse(str_sub(Old_tag, 1, 3) == "OT#", str_sub(Old_tag, 4, -1), Old_tag))

new_master$Old_tag[new_master$Old_tag == "g825"] <- "825"

```


Fix up Old tags (remove # and whitespace)
```{r}

new_master <- mutate(new_master, Old_tag = ifelse(str_sub(Old_tag, 1, 2)== " #",
       str_sub(Old_tag, 3, -1), Old_tag))

new_master <- mutate(new_master, Old_tag = ifelse(str_sub(Old_tag, 1, 1)== " ",
       str_sub(Old_tag, 2, -1), Old_tag))
```


Remove duplicate rows (likely 2014 data)
```{r}
#old and ltreb are both na
both_na <- new_master[(is.na(new_master$Old_tag) & is.na(new_master$LTREB_tag)),]

both_na$Old_tag <- "None"
both_na$LTREB_tag <- "None"
#all rows with at least one tag
removed_na <- new_master[!(is.na(new_master$Old_tag) & is.na(new_master$LTREB_tag)),]


#removed all duplicates, including rows with no tags
deduped <- new_master[!duplicated(new_master[2:8]),]

#removed duplicates, rows with at least one tag
unique_removed <- removed_na[!duplicated(removed_na[2:8]),]

#removed duplicates, rows with no tags
unique_both_na <- both_na[!duplicated(both_na[2:8]),]



#keeps all entries with no tags, with unique at least one tag entries
combined_both_na_unique_removed <- rbind(unique_removed, both_na)
combined_both_na_unique_removed[order(combined_both_na_unique_removed$Sort),]



new_master <- combined_both_na_unique_removed
```

```{r}
u_notes <- unique(new_master$Old_tag)
u_notes <- sort(u_notes, decreasing = TRUE) 
```

```{r}
new_master$Date <- mdy(new_master$Date)
```




**
Error Checks
**

Checking Origin Year vs Stand_year, both are empty (can potentially be extracted from plot name)
```{r}
# oy <- unique(new_master["Origin_year"])
# oy
# sy <- unique(new_master["Stand_year"])
# sy
```

Sanity check that no data was lost 
```{r}
md <- nrow(master_data)
ld <- nrow(live_data)
dd <- nrow(dead_data)
nm <- nrow(new_master)
dp <- nrow(removed_na) - nrow(unique_removed)
sum <- md + ld + dd - dp
sum
nm
glimpse(new_master)
```

Reorganize columns
```{r}
new_master <- new_master %>% select(1,2,14,7,18,21,20,3:6, 8, 12, 9, 10, 19, 16, 17, 13, 15) %>% rename("Outside_historic_plot" = `Outside Historic Plot`)
new_master$Sort <- 1:nrow(new_master)
head(new_master)
```




```{r}
glimpse(new_master)
```

fix agb_kg values

```{r}
new_master$Height_stem <- as.numeric(new_master$Height_stem)

#Generic: COAL, CRRO, OTHER, LALA, PRPE, PRSE, PYSP, RHGL, TSCA, VISP
# height_stem = 2.313 * (DBH_cm)^.562
# AGB_kg = .171 * (DBH_cm)^2.181 * (height_stem)^.4389
generic_codes <- c("COAL", "CRRO", "OTHER", "LALA", "PRPE", "PRSE", "PYSP", "RHGL", "TSCA", "VISP")
new_master <- new_master %>%
                  mutate("Height_stem" = ifelse(Species_code %in% generic_codes, (2.313 * (DBH_cm)^.562), Height_stem)) %>%
                  mutate("AGB_kg" = ifelse(Species_code %in% generic_codes, (.171 * (DBH_cm)^2.181 * (Height_stem)^.4389), AGB_kg) ) %>%
                  mutate(AGB_kg = round(AGB_kg, 1), Height_stem = round(Height_stem, 1))


#FRAM
# height_stem = 3.433 * (DBH_cm)^.577
# AGB_kg = .0185 * (DBH_cm)^1.329 * (height_stem)^1.766
new_master <- new_master %>%
                  mutate("Height_stem" = ifelse(Species_code == "FRAM", (3.433 * (DBH_cm)^.577), Height_stem)) %>%
                  mutate("AGB_kg" = ifelse(Species_code == "FRAM", (.0185 * (DBH_cm)^1.329 * (Height_stem)^1.766), AGB_kg) ) %>%
                  mutate(AGB_kg = round(AGB_kg, 1), Height_stem = round(Height_stem, 1))
glimpse(new_master)


# #TIAM
# # height_stem = 4.9 * (DBH_cm)^.3938
# # AGB_kg = .02322 * (DBH_cm)^2.026 * (height_stem)^.8327
new_master <- new_master %>%
                  mutate("Height_stem" = ifelse(Species_code == "TIAM", (4.9 * (DBH_cm)^.3938), Height_stem)) %>%
                  mutate("AGB_kg" = ifelse(Species_code == "TIAM", (.02322 * (DBH_cm)^2.026 * (Height_stem)^.8327), AGB_kg) ) %>%
                  mutate(AGB_kg = round(AGB_kg, 1), Height_stem = round(Height_stem, 1))
glimpse(new_master)
# 
# #OSVI
# # height_stem = 4.932 * (DBH_cm)^.4183
# # AGB_kg = .06506 * (DBH_cm)^2.269 * (height_stem)^.3964
new_master <- new_master %>% 
                  mutate("Height_stem" = ifelse(Species_code == "OSVI", (4.932 * (DBH_cm)^.4183), Height_stem)) %>%
                  mutate("AGB_kg" = ifelse(Species_code == "OSVI", (.06506 * (DBH_cm)^2.269 * (Height_stem)^.3964), AGB_kg) ) %>%
                  mutate(AGB_kg = round(AGB_kg, 1), Height_stem = round(Height_stem, 1))
glimpse(new_master)
# 
# #SASP
# # height_stem = 2.218 * (DBH_cm)^.5335
# # AGB_kg = .03824 * (DBH_cm)^2.19 * (height_stem)^.5987
new_master <- new_master %>%
                  mutate("Height_stem" = ifelse(Species_code == "SASP", (2.218 * (DBH_cm)^.5335), Height_stem)) %>%
                  mutate("AGB_kg" = ifelse(Species_code == "SASP", (.03824 * (DBH_cm)^2.19 * (Height_stem)^.5987), AGB_kg) ) %>%
                  mutate(AGB_kg = round(AGB_kg, 1), Height_stem = round(Height_stem, 1))
glimpse(new_master)
```

```{r}
plot_names <- unique(new_master$Plot)
plot_names
```


Write merged data to new csv file titled "new_merge_master.csv"
```{r}

```



