---
title: "UMBS LTREB Coarse Woody Debris Table Cleaning"
author: "Alexandria Pawlik"
output: html_notebook
---

Required packages and paths
```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(xlsx)
library(tibble)
library(dataMaid)
options(stringsAsFactors = FALSE)

# set these paths for your machine's specifics
cwd_path_2014 <- "../raw/CWD_2014_raw.csv"
cwd_path_2019 <- "../raw/CWD_2019_raw.csv"
output_path <- "../output/cwd.csv"
```



Read in 2014 data
```{r}
cwd <- read_csv(cwd_path_2014)
head(cwd_data)
species <- sort(unique(cwd_data$`Species code (if possible)`))
species
```

Create status column to represent if stump or not, and take out of species code column
```{r}
cwd_data <- mutate(cwd_data, Status = ifelse(str_sub(`Species code (if possible)`, 1, 8)=="Stump - ", "Stump", NA_character_))

cwd_data <- mutate(cwd_data, `Species code (if possible)` = ifelse(str_sub(`Species code (if possible)`, 1, 8)=="Stump - ", str_sub(`Species code (if possible)`, 9, -1), `Species code (if possible)`))

cwd_data$Status[cwd_data$`Species code (if possible)` == "Stump Piece - PIST"] <- "Stump Piece"
cwd_data$`Species code (if possible)`[cwd_data$`Species code (if possible)` == "Stump Piece - PIST"] <- "PIST"
```

```{r}
species <- sort(unique(cwd_data$`Species code (if possible)`))
species
```

Remove stump from species code + decay class column

```{r}
concat <- sort(unique(cwd_data$`Species Code + Decay Class`))
concat
```

```{r}
cwd_data <- mutate(cwd_data, `Species Code + Decay Class` = ifelse(str_sub(`Species Code + Decay Class`, 1, 8)=="Stump - ", str_sub(`Species Code + Decay Class`, 9, -1), `Species Code + Decay Class`))

cwd_data$`Species Code + Decay Class`[cwd_data$`Species Code + Decay Class` == "Stump Piece - PIST2"] <- "PIST2"
```

```{r}
concat <- sort(unique(cwd_data$`Species Code + Decay Class`))
concat
```

Convert date to standard format
```{r}
cwd_data$`Date Sampled` <- mdy(cwd_data$`Date Sampled`)
```

Reorganize columns
```{r}
cwd_data <- cwd_data %>% select(1:3,5,6, 19, 7:14, 16:18)
```

fix biomass, change 0 to NA
```{r}
#cwd_data change all 0 biomasses to NA
cwd_data$`Estimated Biomass (taper) (g)`[cwd_data$`Estimated Biomass (taper) (g)` == 0] <- NA_character_
```

fix species names, change ? to Unknown
```{r}
cwd_data <- cwd_data %>%  mutate(`Species code (if possible)` = ifelse(str_sub(`Species code (if possible)`, 1, 1) == "?", "Unknown", `Species code (if possible)`))
```

Fix column names to conform to EDI standards
```{r}
cwd_data <- cwd_data %>% rename("Stand_name" = `Stand name`, "Date_sampled" = `Date Sampled`, "Plot_ID" = `Plot ID`, "CWD_subplot_ID" = `CWD subplot ID`, "Species_code" = `Species code (if possible)`, "Length_m" = `Length (m)`, "Endpoint_1_diameter_cm" = `Endpoint 1 diameter (cm)`, "Midpoint_diameter_cm" = `Midpoint diameter (cm)`, "Endpoint_2_diamteter_cm" = `Endpoint 2 diameter (cm)`, "Decay_class_whole_piece" = `Decay class of whole piece`, "Avg_diameter_cm" = `Avg. Diameter (cm)`, "Avg_radius_cm" = `Avg. Radius (cm)`, "Avg_area_cm_squared" = `Avg. Area (cm^2)`, "Estimated_tapered_volume_cm_cubed" = `Est. Tapered Volume (cm^3)`, "Estimated_biomass_tapered_g" = `Estimated Biomass (taper) (g)`)
```


fix plot names to match AGB
```{r}
plot_names <- unique(data.frame(cwd_data$Stand_name, cwd_data$Plot_ID))
plot_names

cwd_data <- cwd_data %>% add_column("Plot" = NA, .before = 1)
cwd_data$Plot = cwd_data$Plot_ID
cwd_data$Plot_ID <- as.character(cwd_data$Plot_ID)

cwd_data$Plot[cwd_data$Plot_ID == "1"] <- "DNR1"
cwd_data$Plot[cwd_data$Plot_ID == "2"] <- "DNR2"
cwd_data$Plot[cwd_data$Plot_ID == "3"] <- "DNR3"
cwd_data$Plot[cwd_data$Plot_ID == "CP#1"] <- "CP1"
cwd_data$Plot[cwd_data$Plot_ID == "CP#2"] <- "CP2"
cwd_data$Plot[cwd_data$Plot_ID == "CP#3"] <- "CP3"
cwd_data$Plot[cwd_data$Plot_ID == "PSA1972#1"] <- "PSA1"
cwd_data$Plot[cwd_data$Plot_ID == "PSA1972#2"] <- "PSA2"
cwd_data$Plot[cwd_data$Plot_ID == "PSA1972#3"] <- "PSA3"
cwd_data$Plot[cwd_data$Plot_ID == "BF1"] <- "BP54#1"
cwd_data$Plot[cwd_data$Plot_ID == "BF2"] <- "BP54#2"

us <- unique(cwd_data$Plot)
us

cwd_data <- cwd_data %>% select(1, 3, 5:18) %>% rename("Estimated_mass_tapered_g" = "Estimated_biomass_tapered_g")
cwd_data <- cwd_data[1:418,]
```



Pull in 2019 data and add date column
```{r}
# AKP
cwd_data <- cwd_data %>% 
  mutate(Length_m = as.numeric(Length_m)) %>% 
  mutate(Estimated_tapered_volume_cm_cubed = as.numeric(Estimated_tapered_volume_cm_cubed)) %>% 
  mutate(Estimated_mass_tapered_g = as.numeric(Estimated_mass_tapered_g))
glimpse(cwd_data)



```

Write csv file
```{r}
write_csv(cwd_data, path = output_path)
```


