---
title: "UMBS LTREB Saplings Table Cleaning"
author: "Alexandria Pawlik"
output: html_notebook
---

Required packages and paths
```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(xlsx)
library(tibble)
library(dataMaid)
options(stringsAsFactors = FALSE)

# set these paths for your machine's specifics
saplings_path_2014 <- "../raw/saplings_2014_raw.csv"
saplings_path_2019 <- "../raw/saplings_2019_raw.csv"

# create a merged table of 2014 and 19, save copy of Alina's version of 2014
output_path_2014 <- "../output/saplings_2014.csv"
output_path_merge <- "../output/saplings.csv"
```


# Create saplings_2014.csv


Read in 2014 data
```{r}
saplings <- read_csv(saplings_path_2014)
saplings <- saplings %>% select(1:16,18:25,27)
head(saplings)
```

Clean up column headers
```{r}
saplings <- saplings %>% 
  rename("DBH_0_to_2" = `0-2`) %>% 
  rename("DBH_0_to_2_Dead" = `0-2, Dead`) %>% 
  rename("DBH_2_to_4" = `2-4`) %>% 
  rename("DBH_2_to_4_Dead" = `2-4, dead`) %>% 
  rename("DBH_4_to_6" = `4-6`) %>% 
  rename("DBH_4_to_6_Dead" = `4-6, dead`) %>% 
  rename("DBH_6_to_8" = `6-8`) %>% 
  rename("DBH_6_to_8_Dead" = `6-8, dead`) %>% 
  rename("Biomass_0_to_2" = `Biomass 0-2`) %>% 
  rename("Biomass_0_to_2_Dead" = `Biomass 0-2, dead`) %>% 
  rename("Biomass_2_to_4" = `Biomass 2-4`) %>% 
  rename("Biomass_2_to_4_Dead" = `Biomass 2-4, dead`) %>% 
  rename("Biomass_4_to_6" = `Biomass 4-6`) %>% 
  rename("Biomass_4_to_6_Dead" = `Biomass 4-6, dead`) %>% 
  rename("Biomass_6_to_8" = `Biomass 6-8`) %>% 
  rename("Biomass_6_to_8_Dead" = `Biomass 6-8, dead`) %>% 
  rename("Mean_Height_0_to_2" =`0-2 Mean Height`) %>% 
  rename("Mean_Height_2_to_4" =`2-4 Mean Height`) %>% 
  rename("Mean_Height_4_to_6" =`4-6 Mean Height`) %>% 
  rename("Mean_Height_6_to_8" =`6-8 Mean Height`) %>% 
  rename("Date" = `Tally Date`)

saplings$"Mean_Height_0_to_2_Dead" = saplings$Mean_height_0_to_2
saplings$"Mean_Height_2_to_4_Dead" = saplings$Mean_height_2_to_4
saplings$"Mean_Height_4_to_6_Dead" = saplings$Mean_height_4_to_6
saplings$"Mean_Height_6_to_8_Dead" = saplings$Mean_height_6_to_8

head(saplings)
```

Fix up tally date column
```{r}
saplings$Date <- mdy(saplings$Date)
head(saplings)
#glimpse(saplings)
```

Merge/gather columns to have "size class" "tally" "biomass" columns, ie wide to long
Done by splitting into two data frames (tally and biomass), gather and reorder separately, and then bind necessary columns
```{r}
saplings_dbh <- saplings %>% select(1:5,7,8,10,11,13,14,15,25)
saplings_bio <- saplings %>% select(1:3,15, 17:25)
saplings_mean <- saplings %>% select(1:3, 6, 21, 9, 22, 12, 23, 16, 24, 15, 25)

saplings_dbh <- saplings_dbh %>% 
  gather(DBH_size_class, Tally, -Plot, -Species, -Date, -`Total Biomass (kg)`, -Index) %>% 
  arrange(Index)

saplings_bio <- saplings_bio %>% 
  gather(DBH_size_class, Biomass, -Plot, -Species, -Date, -`Total Biomass (kg)`, -Index) %>% 
  arrange(Index)

saplings_mean <- saplings_mean %>% 
  gather(DBH_size_class, Mean_Height, -Plot, -Species, -Date, -`Total Biomass (kg)`, -Index) %>% 
  arrange(Index)
```

Edit out "DBH" and "Biomass" from size class column
```{r}
saplings_dbh <- saplings_dbh %>% 
  mutate(DBH_size_class = str_sub(DBH_size_class, 5, -1))

saplings_bio <- saplings_bio %>% 
  mutate(DBH_size_class = str_sub(DBH_size_class, 9, -1))

saplings_mean <- saplings_mean %>% 
  mutate(DBH_size_class = str_sub(DBH_size_class, 13, -1))
```

Merge and reorder
```{r}
saplings_tidy <- cbind(saplings_dbh, Biomass = saplings_bio[[7]], Mean_height = saplings_mean[[7]])

saplings_tidy <- saplings_tidy %>%  
  select(2:4, 6:9, 5)
```

Mutate size class column into size class and status
```{r}
saplings_tidy <- saplings_tidy %>% 
  mutate(Status = ifelse(str_sub(DBH_size_class, -1, -1) =="d", "Dead", "Alive"))

saplings_tidy <- saplings_tidy %>% 
  mutate(DBH_size_class = str_sub(DBH_size_class, 1, 6)) %>% 
  select(1:4, 9, 5:7)
```

Change NA to 0
```{r}
saplings_tidy$Tally[is.na(saplings_tidy$Tally)] <- 0
saplings_tidy$Biomass[saplings_tidy$Tally == 0] <- 0

# check
saplings_tidy %>% filter(is.na(Biomass))
```

Fix plot names to match AGB
```{r}
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#1"] <- "DNR1"
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#2"] <- "DNR2"
saplings_tidy$Plot[saplings_tidy$Plot == "DNR#3"] <- "DNR3"
plot_names <- unique(saplings_tidy$Plot)
plot_names
```

Write output file for 2014 data structure
```{r}
saplings_tidy <- saplings_tidy %>% 
  select(1:7)

# write.csv(saplings_tidy, file = output_path_2014, eol = "\r\n", row.names = FALSE)
```



# Create saplings_merge.csv

Read in 2019 data and prep for merging with 2014
```{r}
saplings_19 = read_csv(saplings_path_2019)
saplings_tidy <- as_tibble(saplings_tidy)

# figure out what needs to be changed for 2019 data to match 2014
glimpse(saplings_19)
glimpse(saplings_tidy)
```

Convert dates
```{r}
saplings_19$Date <- mdy(saplings_19$Date)
head(saplings_19)
```

Convert 2019 wide to long
```{r}
# fix column names to match DBH size classes
saplings_19 <- saplings_19 %>% 
  rename("0_to_2" = `<2 (cm)`) %>% 
  rename("2_to_4" = `2-4 (cm)`) %>% 
  rename("4_to_6" = `4-6 (cm)`) %>% 
  rename("6_to_8" = `6-8 (cm)`)

# wide to long
saplings_19 <- saplings_19 %>% 
  gather(key = "DBH_size_class", value = "Tally", -Plot, -Species, -Date)
```

Change NA to 0
```{r}
saplings_19$Tally[is.na(saplings_19$Tally)] <- 0
```

Fix plot names
```{r}
unique(saplings_19$Plot)

# fix plot names to match AGB
saplings_19$Plot[saplings_19$Plot == "DNR#1"] <- "DNR1"
saplings_19$Plot[saplings_19$Plot == "DNR#2"] <- "DNR2"
saplings_19$Plot[saplings_19$Plot == "DNR#3"] <- "DNR3"
saplings_19$Plot[saplings_19$Plot == "PSA#1"] <- "PSA1"
saplings_19$Plot[saplings_19$Plot == "PSA#2"] <- "PSA2"
saplings_19$Plot[saplings_19$Plot == "PSA#3"] <- "PSA3"

# # add notes column for Q plot name notation
# saplings_19 <- saplings_19 %>% 
#   mutate(Notes = NA)
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q1" | saplings_19$Plot == "BP98#2 Q1"] <- "Q1"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q2" | saplings_19$Plot == "BP98#2 Q2"] <- "Q2"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q3" | saplings_19$Plot == "BP98#2 Q3"] <- "Q3"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q4" | saplings_19$Plot == "BP98#2 Q4"] <- "Q4"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q5" | saplings_19$Plot == "BP98#2 Q5"] <- "Q5"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q6" | saplings_19$Plot == "BP98#2 Q6"] <- "Q6"
# saplings_19$Notes[saplings_19$Plot == "BP98#1 Q7" | saplings_19$Plot == "BP98#2 Q7"] <- "Q7"

# remove Q notation from plot names
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q1"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q2"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q3"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q4"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q5"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q6"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#1 Q7"] <- "BP98#1"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q1"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q2"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q3"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q4"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q5"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q6"] <- "BP98#2"
saplings_19$Plot[saplings_19$Plot == "BP98#2 Q7"] <- "BP98#2"

# combine observations from the same plot
saplings_19 <- saplings_19 %>% 
  group_by(Plot, Species, DBH_size_class, Date) %>% 
  dplyr::summarize(Tally = sum(Tally))
```

Merge lines of same plot different status in 2014 
*live and dead not differentiated in 2019
```{r}
saplings_tidy %>% filter(is.na(Date))
# temporarily change missing dates to make summary easier
saplings_tidy$Date[is.na(saplings_tidy$Date)] <- "3000-01-01"

# collapse dead and alive rows into one
# sum biomass and tally, drop status, merge all others
saplings_14 <- saplings_tidy %>%
  group_by(Plot, Species, DBH_size_class, Date) %>% 
  dplyr::summarize(Tally_sum = sum(Tally), 
            Biomass_sum = sum(Biomass))

# makes dates back into NA
saplings_14$Date[saplings_14$Date == "3000-01-01"] <- NA
```

Combine 2014 and 2019
```{r}
# add empty biomass column to 2019, rearrange columns
saplings_19 <- saplings_19 %>% 
  mutate(Biomass = NA) %>% 
  relocate(Date, .after = 3)

# rename columns
saplings_14 <- saplings_14 %>% 
  ungroup() %>% 
  rename(Tally = Tally_sum) %>% 
  rename(Biomass = Biomass_sum) 

# add Year column for NA dates
saplings_14 <- saplings_14 %>% 
  mutate(Year = "2014", .after = 3)
saplings_19 <- saplings_19 %>% 
  mutate(Year = "2019", .after = 3)

# combine
saplings <- rbind(saplings_14, saplings_19)
saplings <- saplings %>% 
  rename(Plot_ID = Plot) %>% 
  rename(DBH_Size_Class = DBH_size_class) %>% 
  rename(Count = Tally)
glimpse(saplings)
```

Fix species codes
```{r}
saplings$Species[saplings$Species == "AMEL"] <- "AMSP"
saplings$Species[saplings$Species == "Unknown"] <- NA
```

Export to CSV
```{r}
# sort
saplings <- saplings %>% 
  arrange(Plot_ID, Species, DBH_Size_Class, Date)

write.csv(saplings, file = output_path_merge, eol = "\r\n", row.names = FALSE)

# saplings %>% filter(is.na(Plot_ID))
# saplings %>% filter(is.na(Species))
# saplings %>% filter(is.na(DBH_Size_Class))
# saplings %>% filter(is.na(Date))
# saplings %>% filter(is.na(Count))
# saplings %>% filter(is.na(Biomass))

saplings %>% filter(is.na(Date))
```

