---
title: "Master Biomass with 2014 R Notebook"
output: html_notebook
---
NOTE: ALL head(dead_data) OR SIMILAR DEAD DATA CALLS CAUSE PANDOC ERROR 1, WHICH MEANS NOTEBOOK CANNOT BE SAVED WHEN THERE IS HEAD() OUTPUT FOR THIS SPECIFIC DATA SET


This notebook describes the merging process between the Master AGB data and the 2014 inventory. Currently only includes Live and Dead and excludes Sapling


```{r}
library(tidyverse)
library(dataMaid)
library(lubridate)
```

Read in merged master data (pulled from "Master AGB data for UMBS IMS.xlsx" tab "merged master tree list" as a csv file)
```{r}
master_data <- read_csv("../raw/Master AGB data for UMBS IMS merged master tree list.csv")
glimpse(master_data)
```

Read in Live OS data (pulled from "Tree Inventory - Master.xlsx" tab "Live OS Inventory" as a csv file)
```{r}
live_data <- read_csv("../raw/live inventory1.csv")
#glimpse(live_data)
```

Read in Dead OS data (pulled from "Tree Inventory - Master.xlsx" tab "Dead OS Inventory" as a csv file)
```{r}
dead_data <- read_csv("../raw/dead inventory.csv")
# glimpse(dead_data)
```

Mapping between species codes and common names, because there are no NA values for species codes but inconsistent entry for common names, allows for common names to be uniform

**Naming anomalies based on Permanent Plots data  
SASP is technically SALI, for Salix spp.  
VISP is technically VIBU for Viburnum spp.  
PYSP is technically PY for Pyrus  
OTHER implies a species not in the list below was measured  

**How do we want to deal with Unknowns? Is "OTHER" considered unknown?   
```{r}
index <- c("POGR", "PIRE", "PIST", "ACRU", "QURU", "TSCA", "BEPA", "FAGR", "AMSP", "ACSA", "ACPE", "FRAM", "OSVI", "TIAM", "ULAM", "UNKNOWN", "ABBA", "POTR", "THOC", "PIGL", "PRSE", "PRPE", "COAL", "RHGL", "CRRO", "FRNI", "POBA", "ALIN", "LALA", "BEAL", "SASP", "VISP", "PYSP", "PIMA", "OTHER")
values <- c("Populus grandidentata", "Pinus resinosa", "Pinus strobus", "Acer rubrum", "Quercus rubra", "Tsuga canadensis", "Betula papyrifera", "Fagus grandifolia", "Amelanchier", "Acer saccharum", "Acer pensylvanicum", "Fraxinus americana", "Ostrya virginiana", "Tilia americana","Ulmus americana", "UNKNOWN", "Abies balsamea", "Populus tremuloides", "Thuja occidentalis", "Picea glauca", "Prunus serotina", "Prunus pensylvanica", "Cornus alternifolia", "Rhus glabra", "Crataegus roanensis", "Fraxinus nigra", "Populus balsamifera", "Alnus incana", "Larix laricina", "Betula lutea", "Salix spp.", "Viburnum spp.", "Pyrus", "Picea mariana", "OTHER")

live_data <- add_column(live_data, "Common Name" = NA, .after = 8)
live_data$'Common Name' <- values[match(live_data$Species, index)]
glimpse(live_data)
```

```{r}
dead_data <- add_column(dead_data, "Common Name" = NA, .after = 8)
dead_data$'Common Name' <- values[match(dead_data$Species, index)]
# glimpse(dead_data)
```

Sort live and dead data sets so that the columns match better

** Assumptions being made:  
1) Tag # = LTREB tag  
2) Date should be converted to 2014 for Live and Dead (ignore month/day)  
3) DBH (inches) should probably be left blank unless converted to 1in:2.54cm  
4) Biomass estimate (kg) = AGB (kg)  
5) Height (estimate) = ht (does this depend on Alive/Dead status?)  
6) Ignore [Outside historic plot, distance, azimuth, unit circle, unit circle radians, status, notes, plot center x, plot center y, tree x, tree y, random number, volume estimate, biomass estimate (g)]

** To consider:  
- Origin Year and Stand Year are both empty...?  
- do we want to include status?  
- do we want to convert cm to inches for DBH?  
- how do we want to deal with dates? Just assign 2014 to live and dead, ignore m/d?
- are there any more columns that should be included?  
```{r}
live_data_sorted <- live_data 
live_data_sorted <- live_data_sorted[c(3, 9, 8, 13, 1, 11, 20, 12, 2, 4, 5, 6, 7, 10, 14, 15, 16, 17, 18, 19)]
# glimpse(live_data_sorted)
```

```{r}
dead_data_sorted <- dead_data
dead_data_sorted <- dead_data_sorted[c(3, 9, 8, 13, 1, 11, 22, 12, 2, 4, 5, 6, 7, 10, 14, 15, 16, 17, 18, 19, 20, 21)]
#head(dead_data_sorted)
```

Add in missing columns 

```{r}
live_data_sorted <- add_column(live_data_sorted, "Measure Year" = 2014, .after = 4)
live_data_sorted <- add_column(live_data_sorted, "Origin year" = NA, .after = 1)
live_data_sorted <- add_column(live_data_sorted, "Stand Year" = NA, .after = 6)
live_data_sorted <- add_column(live_data_sorted, "DBH (inches)" = NA, .before = 9)
live_data_sorted <- add_column(live_data_sorted, "Sort" = 1, .before = 1) #value 1 denotes live data set
live_data_sorted <- add_column(live_data_sorted, "Old tag" = NA, .before = 6)
live_data_sorted <- add_column(live_data_sorted, "Data Set" = "Live", .after = 14)
glimpse(live_data_sorted)
```

```{r}
dead_data_sorted <- add_column(dead_data_sorted, "Measure Year" = 2014, .after = 4)
dead_data_sorted <- add_column(dead_data_sorted, "Stand Year" = NA, .after = 5)
dead_data_sorted <- add_column(dead_data_sorted, "Origin year" = NA, .after = 1)
dead_data_sorted <- add_column(dead_data_sorted, "DBH (inches)" = NA, .before = 9)
dead_data_sorted <- add_column(dead_data_sorted, "Sort" = 2, .before = 1) #value 2 denotes Dead data set 
dead_data_sorted <- add_column(dead_data_sorted, "Old tag" = NA, .before = 6)
dead_data_sorted <- add_column(dead_data_sorted, "Data Set" = "Dead", .after = 14)
#head(dead_data_sorted)

```


Temporary data sets that will be combined into the master set (I currently am only using columns that are in the master set, ignoring all others. Not sure what new columns to put in master)
```{r}
live_to_combine <- live_data_sorted
live_to_combine <- live_to_combine[c(1:9, 11:15)]
live_to_combine <- live_to_combine %>% rename("old tag" = `Old tag`, "Species code" = Species, "LTREB tag" = `Tag #`, "dbh (inches)" = `DBH (inches)`, "dbh (cm)" = `DBH (cm)`, "AGB (kg)" = `Biomass estimate (kg)`, "ht" = `Height (estimate)`)
# glimpse(live_to_combine)
#glimpse(live_to_combine)
```

```{r}
dead_to_combine <- dead_data_sorted
dead_to_combine <- dead_to_combine[c(1:9, 11:15)]
dead_to_combine <- dead_to_combine %>% rename("old tag" = `Old tag`, "Species code" = Species, "LTREB tag" = `Tag #`, "dbh (inches)" = `DBH (inches)`, "dbh (cm)" = `DBH (cm)`, "AGB (kg)" = `Biomass estimate (kg)`, "ht" = `Height (m) estimate`)
#head(dead_to_combine)
```

```{r}
master_to_combine <- master_data
master_to_combine <- add_column(master_to_combine, "Data Set" = "Master", .after = 14)
```

Combine data sets
```{r}
#m <- names(master_data)
#l <- names(live_to_combine)
#d <- names(dead_to_combine)

#setdiff(m, l)
#setdiff(m, d)
#setdiff(l,d)
#new_master <- rbind(master_data, live_to_combine, dead_to_combine)
new_master <- rbind(master_to_combine, live_to_combine, dead_to_combine)
#glimpse(new_master)
```


Using species code to common names mapping, fixes the majority of common name inconsistencies (any species codes not in the mapping will have NA common name)
```{r}
new_master$"Species code"[new_master$"Species code" == "UNKOWN"] <- "UNKNOWN" #fix misspellings
new_master$"Species code"[new_master$"Species code" == "Coal"] <- "COAL"
sc <- unique(new_master["Species code"])
sc
new_master$'Common Name' <- values[match(new_master$"Species code", index)]
cns <- unique(new_master["Common Name"])
cns

head(new_master)
```


Add sort values for live and dead data
```{r}
new_master$"Sort" <- c(1:nrow(new_master))
tail(new_master)
```

Clean up column titles
```{r}
new_master <- rename(new_master, "Origin Year" = `Origin year`, "Old Tag" = `old tag`, "LTREB Tag" = `LTREB tag`, "DBH (inches)" = `dbh (inches)`, "DBH (cm)" = `dbh (cm)`, "Height (m)" = `ht`)
head(new_master)
# glimpse(new_master)
```

Filling in DBH (inches) 
```{r}
inch <- new_master %>% filter(is.na(new_master$"DBH (inches)"))
inch
#new_master$"DBH (inches)"[is.na(new_master$"DBH (inches)")] <- round((new_master$"DBH (cm)")/2.56, 1)

```



Checking for Species codes with no corresponding Common Name
```{r}
temp <- new_master %>% filter(is.na(new_master$"Common Name"), !is.na(new_master$"Species code"))
#glimpse(temp)
sc <- unique(temp["Species code"])
sc
```

Get unique plots
```{r}
plot_unique <- unique(new_master["Plot"])
plot_unique
```

Sanity check comparing original data and merged data 
```{r}

fifty_yr_5 <- read_csv("../raw/50yr5.csv")
names(fifty_yr_5) <- as.character(unlist(fifty_yr_5[3,]))
fifty_yr_5[-c(1,2,3),]

fy5_master_potr <-  new_master %>% filter(new_master$"Species code" == "POTR", new_master$"Plot" == "50YR#5", new_master$"Measure Year" == "1951")
fy5_orig_potr <- fifty_yr_5 %>% filter(fifty_yr_5$"SPECIES" == "27", !is.na(fifty_yr_5$"1951"))

fy5_master_potr
fy5_orig_potr

fy5_master_acru <- new_master %>% filter(new_master$"Species code" == "ACRU", new_master$"Plot" == "50YR#5", new_master$"Measure Year" == "1951")
fy5_orig_acru <- fifty_yr_5 %>% filter(fifty_yr_5$"SPECIES" == "1", !is.na(fifty_yr_5$"1951"))

fy5_master_acru
fy5_orig_acru

```

Checking Origin Year vs Stand Year, both are empty? 
```{r}
oy <- unique(new_master["Origin Year"])
oy
sy <- unique(new_master["Stand Year"])
sy
```

Checking tags, wondering if possible to combine somehow
```{r}
no_tags <- new_master %>% filter(is.na(new_master$"Old Tag"), is.na(new_master$"LTREB Tag"))
no_tags #21,613 don't have any tag number
nrow(new_master) #44,345 entries total
no_ltreb <- new_master %>% filter(!is.na(new_master$"LTREB Tag"))
no_ltreb #4,242 have LTREB tag number
no_old <-new_master %>% filter(!is.na(new_master$"Old Tag"))
no_old #18,623 have old tag number
```

Sanity check that no data was lost 
```{r}
md <- nrow(master_data)
ld <- nrow(live_data)
dd <- nrow(dead_data)
nm <- nrow(new_master)
sum <- md + ld + dd
sum
nm
```

```{r}
bad_names <- c("Old Tag ", "Old tag ")
new_master$"Old Tag" <- ifelse(substr(new_master$"Old Tag", 1, 8)  %in% bad_names, substr(new_master$"Old Tag", 8, str_length(new_master$"Old Tag")), new_master$"Old Tag")
new_master$"Old Tag" <- ifelse(substr(new_master$"Old Tag", 1, 7)  == "Old tag", substr(new_master$"Old Tag", 7, str_length(new_master$"Old Tag")), new_master$"Old Tag")
```

```{r}
new_master <- add_column(new_master, "As Class" = NA, .after = 11)
```


```{r}
similar <- c("50YR#2", "50YR#3", "50YR#4", "50YR#5")
new_master$"As Class"[new_master$"Measure Year" <= 1979 & new_master$"Plot" %in% similar] <- "Yes"
new_master$"As Class"[new_master$"Plot" == "50YR#6"] <- "Yes"
new_master$"As Class"[new_master$"Measure Year" <= 1981 & new_master$"Plot" == "50YR#7"] <- "Yes"
head(new_master)
tail(new_master)
```



```{r}
write_csv(new_master, path = "../output/TEST_Biomass_w_2014.csv")
```


